Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-05-15 talk/dls2014/paper/paper.tex
paper updates

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages (e.g.
PyPy, V8, Tracemonkey) started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g. Python and Ruby use a single,
global interpreter lock (GIL) to serialize the execution of code in
threads.

While this GIL prevents any parallelism from occurring, it also
provides some useful guarantees. Since this lock is always acquired
while executing bytecode instructions and it may only be released
in-between such instructions, it provides perfect isolation and
atomicity between multiple threads for a series of
instructions. Another technology that can provide the same guarantees
is transactional memory (TM).

There have been several attempts at replacing the GIL with TM. Using
transactions to enclose multiple bytecode instructions, we can get the
very same semantics as the GIL while possibly executing several
transactions in parallel. Furthermore, by exposing these
interpreter-level transactions to the application in the form of
\emph{atomic blocks}, we give dynamic languages a new synchronization
mechanism that avoids several of the problems of locks as they are
used now.

\cfbolz{the above is good, here is something missing: problems with current STM approaches, outlining the intuition behind the new one}

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of hardware features and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in Python and is
  then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronization mechanism for
  threads.
\end{itemize}

@@

Transactional memory (TM) is a concurrency control mechanism that
comes from database systems. Using transactions, we can group a series
of instructions performing operations on memory and make them happen
atomically and in complete isolations from other
transactions. \emph{Atomicity} means that all these instructions in
the transaction and their effects seem to happen at one, undividable
point in time. Other transactions never see inconsistent state of a
partially executed transaction which is called \emph{isolation}.

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serializable}. Meaning, the outcome is equal to some sequential
execution of these transactions. This means that the approach provides the same
semantics as using the GIL
while still allowing the TM system to
run transactions in parallel as an optimization.

@@

\subsection{Synchronization}

@@

It is well known that using locks to synchronize multiple threads is
hard. They are non-composable, have overhead, may deadlock, limit
scalability, and overall add a lot of complexity. For a better
parallel programming model for dynamic languages, we want to implement
another, well-known synchronization mechanism: \emph{atomic blocks}.

@@

In this section, we describe the general model of our TM system. This
should clarify the general semantics in commonly used terms from
the literature.

\cfbolz{there is an overview paragraph of the idea missing, maybe in the introduction}

\cfbolz{this all feels very much dumping details, needs more overview. why is this info important? the subsubsections don't have any connections}

@@

Our conflict detection works with \emph{object
  granularity}. Conceptually, it is based on \emph{read} and
\emph{write sets} of transactions.  Two transactions conflict if they
have accessed a common object that is now in the write set of at least
one of them.

The \emph{concurrency control} works partly \emph{optimistically} for
reading of objects, where conflicts caused by just reading an object
in transactions are detected only when the transaction that writes the
object actually commits. For write-write conflicts we are currently
\emph{pessimistic}: Only one transaction may have a certain object in
its write set at any point in time, others trying to write to it will
have to wait or abort.

We use \emph{lazy version management} to ensure that modifications by
a transaction are not visible to another transaction before the former
commits.

@@

As required for TM systems, we guarantee complete \emph{isolation}
and \emph{atomicity} for transactions at all times. Furthermore,
the isolation provides full \emph{opacity} to always guarantee a consistent
read set.

To support irreversible operations that cannot be undone when we abort
a transaction (e.g. I/O, syscalls, and non-transactional code in
general), we employ \emph{irrevocable} or \emph{inevitable
transactions}. These transactions are always guaranteed to
commit. There is always at most one such transaction running in the
system, thus their execution is serialised. With this guarantee,
providing \emph{strong isolation} and \emph{serializability} between
non-transactional code is possible by making the current transaction
inevitable right before running irreversible operations.

\subsubsection{Contention Management}

When a conflict is detected, we perform some simple contention
management.  First, inevitable transactions always win. Second, the
older transaction wins. Different schemes are possible.

\subsubsection{Software Transactional Memory}

Generally speaking, the system is fully implemented in
software. However, we exploit some more advanced features of current
CPUs, especially \emph{memory segmentation, virtual memory,} and the
64-bit address space.

@@

During runtime, we can then privatize single pages in segments $>0$
again by remapping single pages as seen in (III).

Looking back at address translation for object references, we see now
that this is actually a two-step process. First, $\%gs{::}SO$ gets
translated to different linear addresses in different threads by the
CPU. Then, depending on the current mapping of virtual pages to file
pages, these LAs can map to a single file page in the sharing-segment,
or to privatized file pages in the corresponding segments. This
mapping is also performed efficiently by the CPU and can easily be
done on every access to an object.

@@

    \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
      segment 0, fully shared memory configuration. (III) privatize single
      pages.\label{fig:Page-Remapping}}
\end{figure}

@@

To detect when to privatize pages, we use write barriers before every
write. When the barrier detects that the object is not in a private
page (or any pages that belong to the object), we remap and copy the
pages to the thread's segment. From now on, the translation of
$\%gs{::}SO$ in this particular segment will resolve to the private
version of the object. Note, the SO used to reference the object does
not change during that process.

@@

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.\\
  Re-sharing these pages instead of resetting the changes would not
  only be slower because it involves a system call, it is also very
  likely the page would get privatised again immediately after re-trying
  the transaction.  Since this again involves a full page copy,
  resetting should be faster than re-sharing.
\end{description}

\cfbolz{random question: did we investigate the extra memory requirements? we should characterize memory overhead somewhere, eg at least one byte per object for the read markers}

@@

We provide isolation between transactions by privatizing the pages of
the segments belonging to the threads the transactions run in.  To
detect when and which pages need privatization, we use write barriers
that trigger a COW of one or several pages. Conflicts, however, are
detected on the level of objects; based on the concept of read and
write sets. Barriers before reading and writing add objects to the
corresponding set; particularly detecting write-write conflicts
eagerly.  On commit, we resolve read-write conflicts and push
modifications to other segments. Aborting transactions simply undo
their changes by copying from the sharing-segment.

@@

However, the layout of a segment is not uniform and we actually
privatize a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if SO is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

To add the object to the read set, for us it is enough to mark it as
read. Since this information needs to be local to the segment, we need
to store it in private pages. The area is called \emph{read markers
}and already mentioned in section \ref{sub:Setup}. This area can be
seen as a continuous array of bytes that is indexed from the start of
the segment by an object's reference ($SO$) divided by 16 (this
requires objects of at least 16 bytes in size). Instead of just
setting the byte to \lstinline!true!  if the corresponding object was
read, we set it to a \lstinline!read_version!  belonging to the
transaction, which will be incremented on each commit.  Thereby, we
can avoid resetting the bytes to \lstinline!false!  on commit and only
need to do this every 255 transactions. The whole code for the barrier
is easily optimizable for compilers as well as perfectly predictable
for CPUs:

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_read(SO):
    *(SO >> 4) = read_version
\end{lstlisting}

@@

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void stm_write(SO):
	if SO->flags & WRITE_BARRIER:
		write_slowpath(SO)
\end{lstlisting}

@@

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
void write_slowpath(SO):
	// GC part:
	list_append(to_trace, SO)
	if is_overflow_obj(SO):
		SO->flags &= ~WRITE_BARRIER
		return
	// STM part
	stm_read(SO)
	lock_idx = SO >> 4
  retry:
	if write_locks[lock_idx] == our_num:
		// we already own it
	else if write_locks[lock_idx] == 0:
		if cmp_and_swap(&write_locks[lock_idx],
					    0, our_num):
			list_append(modified_old_objects, SO)
			privatize_pages(SO)
		else:
			goto retry
	else:
		w_w_contention_management()
		goto retry
	SO->flags &= ~WRITE_BARRIER
\end{lstlisting}

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to. The check for \lstinline!is_overflow_obj()!  tells us if the
object was actually created in this transaction. In that case, we do
not need to execute the following \emph{TM part}.  We especially do
not need to privatize the page since no other transaction knows about
these ``old'' objects.

For TM, we first perform a read barrier on the object. We then try to
acquire its write lock. \lstinline!write_locks!  again is a simple
global array of bytes that is indexed with the SO of the object
divided by 16. If we already own the lock, we are done.  If someone
else owns the lock, we will do a write-write contention management
that will abort either us or the current owner of the object.  If we
succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatize
all pages belonging to it (copy-on-write).

@@

Aborting a transaction is rather easy. The first step is to reset the
nursery and all associated data structures. The second step is to go
over all objects in the write set (\lstinline!modified_old_objects!)
and reset any modifications in our private pages by copying from the
sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
back to the location initialized by a \lstinline!setjmp()!  in
\lstinline!stm_start_transaction()!.  Increasing the
\lstinline!read_version!  is also done there.

@@

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).

@@

\subsubsection{Thread Synchronization}

A requirement for performing a commit is to synchronize all threads so
that we can safely update objects in other segments. To make this
synchronization fast and cheap, we do not want to insert an additional
check regularly in order to see if synchronization is requested. We
use a trick relying on the fact that dynamic languages are usually
very high-level and thus allocate a lot of objects very regularly.
This is done through the function \lstinline!stm_allocate!  shown
below:

\begin{lstlisting}[basicstyle={\footnotesize\ttfamily},tabsize=4]
object_t *stm_allocate(ssize_t size_rounded):
    result = nursery_current
	nursery_current += size_rounded
	if nursery_current > nursery_end:
		return allocate_slowpath(size_rounded)
	return result
\end{lstlisting}

@@

If we want to synchronize all threads, we can rely on this check being
performed regularly. So what we do is to set the
\lstinline!nursery_end!  to $0$ in all segments that we want to
synchronize. The mentioned check will then fail in those segments and
call the slow path. In \lstinline!allocate_slowpath!  they can simply
check for this condition and enter a safe point.

For other synchronization requirements, for example:
\begin{itemize}[noitemsep]
\item waiting for a segment to be released,
\item waiting for a transaction to abort or commit,
\item waiting for all threads to reach their safe points,
\end{itemize}
we use a set of condition variables to wait or signal other threads.

>>>>

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages (e.g.
PyPy, V8, Tracemonkey) started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g. Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads.

While this GIL prevents any parallelism from occurring, it also
provides some useful guarantees. Since this lock is always acquired
while executing bytecode instructions and it may only be released
in-between such instructions, it provides perfect isolation and
atomicity between multiple threads for a series of
instructions. Another technology that can provide the same guarantees
is transactional memory (TM). \remi{cite our position paper}

There have been several attempts at replacing the GIL with TM. Using
transactions to enclose multiple bytecode instructions, we can get the
very same semantics as the GIL while possibly executing several
transactions in parallel. Furthermore, by exposing these
interpreter-level transactions to the application in the form of
\emph{atomic blocks}, we give dynamic languages a new synchronisation
mechanism that avoids several of the problems of locks as they are
used now.

\remi{cite and extract from (our pos. paper):}
TM systems come in can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints, while STM systems have a lot of overhead.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of hardware features and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in Python and is
  then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

Transactional memory (TM) is a concurrency control mechanism that
comes from database systems. Using transactions, we can group a series
of instructions performing operations on memory and make them happen
atomically and in complete isolation from other
transactions. \emph{Atomicity} means that all these instructions in
the transaction and their effects seem to happen at one, indivisible
point in time. Other transactions never see inconsistent state of a
partially executed transaction which is called \emph{isolation}.

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serialisable}. Meaning, the outcome is equal to some sequential
execution of these transactions. This means that the approach provides the same
semantics as using the GIL
while still allowing the TM system to
run transactions in parallel as an optimisation.

@@

\subsection{Synchronisation}

@@

It is well known that using locks to synchronise multiple threads is
hard. They are non-composable, have overhead, may deadlock, limit
scalability, and overall add a lot of complexity. For a better
parallel programming model for dynamic languages, we want to implement
another, well-known synchronisation mechanism: \emph{atomic blocks}.

@@

In this section, we characterise the model of our TM system and its
guarantees as well as some of the design choices we made. This should
clarify the general semantics in commonly used terms from the
literature.\remi{cite Transactional Memory 2nd edition}

Our TM system is fully implemented in software. However, we do exploit
some more advanced features of current CPUs, particularly \emph{memory
segmentation, virtual memory,} and the 64-bit address space. Still,
it cannot be classified as a hybrid TM system since it currently
makes no use of any HTM present in the CPU.

@@

We implement an object-based TM system, thus it makes sense to detect
conflicts with \emph{object granularity}. With this choice, if two
transactions access the same object and at least one access is a
write, we count it as a conflict. Conceptually, it is based on
\emph{read} and \emph{write sets} of transactions. Reading from an
object adds the object to the read set, writing to it adds it to both
sets. Two transactions conflict if they have accessed a common object
that is in the write set of at least one of them.

The detection, or \emph{concurrency control}, works partly
\emph{optimistically} for reading objects. Read-write conflicts
between two transactions are detected in both exactly at the time when
the writing one commits. For write-write conflicts we are currently
\emph{pessimistic}: Only one transaction may have a certain object in
its write set at any point in time, others trying to write to it will
have to wait or abort. This decision needs to be evaluated further
in the future.

When a conflict is detected, we perform some simple contention
management that generally prefers the older transaction to the younger.

@@

As required for TM systems, we guarantee complete \emph{isolation} and
\emph{atomicity} for transactions at all times. Our method of choice
is \emph{lazy version management}. Modifications by a transaction are
not visible to another transaction before the former commits.
Furthermore, the isolation provides full \emph{opacity} to always
guarantee a consistent read set even for non-committed transactions.
\remi{cite On the Correctness of Transactional Memory}

To also support these properties for irreversible operations that
cannot be undone when we abort a transaction (e.g. I/O, syscalls, and
non-transactional code in general), we use \emph{irrevocable} or
\emph{inevitable transactions}. These transactions are always
guaranteed to commit, which is why they always have to win in case
there is a conflict with another, normal transaction. There is always
at most one such transaction running in the system, thus their
execution is serialised. With this guarantee, providing \emph{strong
isolation} and \emph{serialisability} between non-transactional code
is possible by making the current transaction inevitable right before
running irreversible operations.

@@

During runtime, we can then privatise single pages in segments $>0$
again by remapping single pages as seen in (III).

Looking back at address translation for object references, we see now
that this is actually a two-step process. First, $\%gs{::}SO$ gets
translated to different linear addresses in different threads by the
CPU. Then, depending on the current mapping of virtual pages to file
pages, these LAs can map to a single file page in the sharing-segment,
or to privatised file pages in the corresponding segments. This
mapping is also performed efficiently by the CPU and can easily be
done on every access to an object.

@@

    \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
      segment 0, fully shared memory configuration. (III) privatise single
      pages.\label{fig:Page-Remapping}}
\end{figure}

@@

To detect when to privatise pages, we use write barriers before every
write. When the barrier detects that the object is not in a private
page (or any pages that belong to the object), we remap and copy the
pages to the thread's segment. From now on, the translation of
$\%gs{::}SO$ in this particular segment will resolve to the private
version of the object. Note, the SO used to reference the object does
not change during that process.

@@

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronise all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.\\
  Re-sharing these pages instead of resetting the changes would not
  only be slower because it involves a system call, it is also very
  likely the page would get privatised again immediately after re-trying
  the transaction.  Since this again involves a full page copy,
  resetting should be faster than re-sharing.
\end{description}

\cfbolz{random question: did we investigate the extra memory requirements? we should characterise memory overhead somewhere, eg at least one byte per object for the read markers}

@@

We provide isolation between transactions by privatising the pages of
the segments belonging to the threads the transactions run in.  To
detect when and which pages need privatisation, we use write barriers
that trigger a COW of one or several pages. Conflicts, however, are
detected on the level of objects; based on the concept of read and
write sets. Barriers before reading and writing add objects to the
corresponding set; particularly detecting write-write conflicts
eagerly.  On commit, we resolve read-write conflicts and push
modifications to other segments. Aborting transactions simply undo
their changes by copying from the sharing-segment.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if SO is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

To add the object to the read set, for us it is enough to mark it as
read. Since this information needs to be local to the segment, we need
to store it in private pages. The area is called \emph{read markers
}and already mentioned in section \ref{sub:Setup}. This area can be
seen as a continuous array of bytes that is indexed from the start of
the segment by an object's reference ($SO$) divided by 16 (this
requires objects of at least 16 bytes in size). Instead of just
setting the byte to \lstinline!true!  if the corresponding object was
read, we set it to a \lstinline!read_version!  belonging to the
transaction, which will be incremented on each commit.  Thereby, we
can avoid resetting the bytes to \lstinline!false!  on commit and only
need to do this every 255 transactions. The whole code for the barrier
is easily optimisable for compilers as well as perfectly predictable
for CPUs:

\begin{lstlisting}
void stm_read(SO):
    *(SO >> 4) = read_version
\end{lstlisting}

@@

\begin{lstlisting}
void stm_write(SO):
	if SO->flags & WRITE_BARRIER:
		write_slowpath(SO)
\end{lstlisting}

@@

\begin{lstlisting}
void write_slowpath(SO):
	// GC part:
	list_append(to_trace, SO)
	if is_overflow_obj(SO):
		SO->flags &= ~WRITE_BARRIER
		return
	// STM part
	stm_read(SO)
	lock_idx = SO >> 4
  retry:
	if write_locks[lock_idx] == our_num:
		// we already own it
	else if write_locks[lock_idx] == 0:
		if cmp_and_swap(&write_locks[lock_idx],
					    0, our_num):
			list_append(modified_old_objects, SO)
			privatize_pages(SO)
		else:
			goto retry
	else:
		w_w_contention_management()
		goto retry
	SO->flags &= ~WRITE_BARRIER
\end{lstlisting}

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to. The check for \lstinline!is_overflow_obj()!  tells us if the
object was actually created in this transaction. In that case, we do
not need to execute the following \emph{TM part}.  We especially do
not need to privatise the page since no other transaction knows about
these ``old'' objects.

For TM, we first perform a read barrier on the object. We then try to
acquire its write lock. \lstinline!write_locks!  again is a simple
global array of bytes that is indexed with the SO of the object
divided by 16. If we already own the lock, we are done.  If someone
else owns the lock, we will do a write-write contention management
that will abort either us or the current owner of the object.  If we
succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

Aborting a transaction is rather easy. The first step is to reset the
nursery and all associated data structures. The second step is to go
over all objects in the write set (\lstinline!modified_old_objects!)
and reset any modifications in our private pages by copying from the
sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
back to the location initialised by a \lstinline!setjmp()!  in
\lstinline!stm_start_transaction()!.  Increasing the
\lstinline!read_version!  is also done there.

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).

@@

\subsubsection{Thread Synchronisation}

A requirement for performing a commit is to synchronise all threads so
that we can safely update objects in other segments. To make this
synchronisation fast and cheap, we do not want to insert an additional
check regularly in order to see if synchronisation is requested. We
use a trick relying on the fact that dynamic languages are usually
very high-level and thus allocate a lot of objects very regularly.
This is done through the function \lstinline!stm_allocate!  shown
below:

\begin{lstlisting}
object_t *stm_allocate(ssize_t size_rounded):
    result = nursery_current
	nursery_current += size_rounded
	if nursery_current > nursery_end:
		return allocate_slowpath(size_rounded)
	return result
\end{lstlisting}

@@

If we want to synchronise all threads, we can rely on this check being
performed regularly. So what we do is to set the
\lstinline!nursery_end!  to $0$ in all segments that we want to
synchronise. The mentioned check will then fail in those segments and
call the slow path. In \lstinline!allocate_slowpath!  they can simply
check for this condition and enter a safe point.

For other synchronisation requirements, for example:
\begin{itemize}[noitemsep]
\item waiting for a segment to be released,
\item waiting for a transaction to abort or commit,
\item waiting for all threads to reach their safe points,
\end{itemize}
we use a set of condition variables to wait or signal other threads.
